---
output:
  html_document: default
  pdf_document: default
---
Explore and Summarize Data: Red Wine Dataset 
by John Ortiz
========================================================

```{r echo=FALSE, message=FALSE, warning=FALSE, packages}
# Load all of the packages that you end up using in your analysis in this code
# chunk.
library(memisc)
library(ggplot2)
library(gridExtra)
library(GGally)
library(corrplot)
library(dplyr)
library(pander)

```
## Introduction

For this project, I will analyze the red wine dataset. I will try to determine 
what values of each variable make for the quailty of the wine. First I will 
perform a univariate, bivariate, and multivariate analysis. 

>**Dataset Background**: This tidy data set contains 1,599 red wines with 11 
variables on the chemical properties of the wine. At least 3 wine experts rated 
the quality of each wine, providing a rating between 0 (very bad) and 
10 (very excellent). 

>The dataset is related to a red variant of the Portuguese "Vinho Verde" wine. 
For more details, consult: http://www.vinhoverde.pt/en/ or the reference 
[Cortez et al., 2009]. Due to privacy and logistic issues, only physicochemical 
(inputs) and sensory (the output) variables are available (e.g. there is no data
about grape types, wine brand, wine selling price, etc.).

>These datasets can be viewed as classification or regression tasks. The classes
are ordered and not balanced (e.g. there are munch more normal wines than 
excellent or poor ones). Outlier detection algorithms could be used to detect 
the few excellent or poor wines. Also, we are not sure if all input variables 
are relevant. So it could be interesting to test feature selection methods. 

```{r }

# Load the Data
wine <- read.csv('wineQualityReds.csv')

```


## Structure and summary of the Dataframe

```{r echo=FALSE, message=FALSE, warning=FALSE}
# Check dataset structure
str(wine)
```

```{r echo=FALSE, message=FALSE, warning=FALSE}
# Compute summary statistics
summary(wine)
```

Let's do some modifications to the dataset. Quality seems like it would best fit
as a ordered factor. Let's us also add a new ordered factored variable called 
'rating' based on the quailty of the wine.

```{r}
#Convert Quality to a Factor
wine$quality <- factor(wine$quality, ordered = T)
#Create 'Rating' as an ordered factor
wine$rating <- ifelse(wine$quality < 5, 'bad', ifelse(
  wine$quality < 7, 'average', 'good'))
wine$rating <- ordered(wine$rating,
                       levels = c('bad', 'average', 'good'))
```


## Univariate Plots

Let's do a plot the distribution of each of the variable. Also, there are 
extreme outliers for many variables, let's remove them so a better analysis can 
be performed.

```{r echo=FALSE, message=FALSE, warning=FALSE}
#plot barchart for quality
ggplot(data = wine, aes(x = quality)) +
  geom_bar(width = 1,fill = I('red'),color = 'black')
```

```{r echo=FALSE, message=FALSE, warning=FALSE}
#plot barchart for rating
ggplot(data = wine, aes(x = rating)) +
  geom_bar(width = 1,fill = I('red'),color = 'black')
```

**INSIGHT:** Most of the wines present in the dataset are average quality wines.
```{r echo=FALSE, message=FALSE, warning=FALSE}
#plot boxplot for fixed.acidity
ggplot(wine, aes( x = 1, y = fixed.acidity ) ) + 
               geom_jitter(alpha = 0.1 ) +
               geom_boxplot(alpha = 0.2, color = 'red' ) +
               scale_y_continuous(lim = c(4,14))
```
```{r echo=FALSE, message=FALSE, warning=FALSE}
#plot barchart for fixed.acidity
ggplot(data = wine, aes(x = fixed.acidity)) +
  geom_histogram(binwidth = 1, color = 'black',fill = I('red')) + 
  scale_x_continuous(lim = c(4,14))
```

**INSIGHT:** Fixed Acidity is positively skewed. The median is around 8ish.

```{r echo=FALSE, message=FALSE, warning=FALSE}
#plot boxplot for volatile.acidity
ggplot(wine, aes( x = 1, y = volatile.acidity ) ) + 
               geom_jitter(alpha = 0.1 ) +
               geom_boxplot(alpha = 0.2, color = 'red' ) +
               scale_y_continuous(lim = c(0,1))
```

```{r echo=FALSE, message=FALSE, warning=FALSE}
#plot barchart for volatile.acidity
ggplot(data = wine, aes(x = volatile.acidity)) +
  geom_histogram(binwidth = 0.05, color = 'black',fill = I('red')) +
  scale_x_continuous(lim = c(0,1))
```

**INSIGHT:** There seems to be a Bimodal distribution for Volatile acidity, with
peaks at around 0.375 and 0.625.

```{r echo=FALSE, message=FALSE, warning=FALSE}
#plot boxplot for citric.acid
ggplot(wine, aes( x = 1, y = citric.acid )) + 
               geom_jitter(alpha = 0.1 ) +
               geom_boxplot(alpha = 0.2, color = 'red' )
```

```{r echo=FALSE, message=FALSE, warning=FALSE}
#plot barchart for citric.acid
ggplot(data = wine, aes(x = citric.acid)) +
  geom_histogram(binwidth = 0.08, color = 'black',fill = I('red')) +
  scale_x_continuous(breaks = seq(0,1,0.1), lim = c(0,1))
```

**INSIGHT:**The distribution of Citric acid looks a bit off. The seem to be
missing data in the higher values. Maybe the data collected was incomplete?

```{r echo=FALSE, message=FALSE, warning=FALSE}
#plot boxplot for residual.sugar
ggplot(wine, aes( x = 1, y = residual.sugar )) + 
               geom_jitter(alpha = 0.1 ) +
               geom_boxplot(alpha = 0.2, color = 'red' ) +
               scale_y_continuous(lim = c(1,8))
```
```{r echo=FALSE, message=FALSE, warning=FALSE}
#plot barchart for residual.sugar
ggplot(data = wine, aes(x = residual.sugar)) +
  geom_histogram(binwidth = 0.1, color = 'black',fill = I('red')) +
  scale_x_continuous(lim = c(1,8))
```

**INSIGHT:** Residual Sugar is positively skewed with high peaks at around 2.2. 
There are a lot of outliers present at the higher values.

```{r echo=FALSE, message=FALSE, warning=FALSE}
#plot boxplot for chlorides
ggplot(wine, aes( x = 1, y = chlorides )) + 
               geom_jitter(alpha = 0.1 ) +
               geom_boxplot(alpha = 0.2, color = 'red' ) +
               scale_y_continuous(lim = c(0,0.25))
```
```{r echo=FALSE, message=FALSE, warning=FALSE}
#plot barchart for chlorides
ggplot(data = wine, aes(x = chlorides)) +
  geom_histogram(binwidth = 0.01, color = 'black',fill = I('red')) +
  scale_x_continuous(lim = c(0,0.25))
```


**INSIGHT:** Chlorides is positively skewed with high peaks at around 0.25.

```{r echo=FALSE, message=FALSE, warning=FALSE}
#plot boxplot for free.sulfur.dioxide
ggplot(wine, aes( x = 1, y = free.sulfur.dioxide )) + 
               geom_jitter(alpha = 0.1 ) +
               geom_boxplot(alpha = 0.2, color = 'red' ) +
               scale_y_continuous(lim = c(0,45))
```

```{r echo=FALSE, message=FALSE, warning=FALSE}
#plot barchart for free.sulfur.dioxide
ggplot(data = wine, aes(x = free.sulfur.dioxide)) +
  geom_histogram(binwidth = 1, color = 'black',fill = I('red')) +
  scale_x_continuous(breaks = seq(0,80,5), lim = c(0,45))
```


**INSIGHT:** Free Sulphur Dioxide has a high peak at around 5. It is also 
positively skewed with a long tail. Some outliers in the high values are present

```{r echo=FALSE, message=FALSE, warning=FALSE}
#plot boxplot for total.sulfur.dioxide
ggplot(wine, aes( x = 1, y = total.sulfur.dioxide )) + 
               geom_jitter(alpha = 0.1 ) +
               geom_boxplot(alpha = 0.2, color = 'red' ) +
               scale_y_continuous(lim = c(0,180))
```

```{r echo=FALSE, message=FALSE, warning=FALSE}
#plot barchart for total.sulfur.dioxide
ggplot(data = wine, aes(x = total.sulfur.dioxide)) +
  geom_histogram(binwidth = 5, color = 'black',fill = I('red')) +
  scale_x_continuous(lim = c(0,180))
```

**INSIGHT:** Total Sulphur Dioxide has a high peak at around 25. It is also 
positively skewed with a long tail. Some outliers in the high values are present

```{r echo=FALSE, message=FALSE, warning=FALSE}
#plot boxplot for density
ggplot(wine, aes( x = 1, y = density)) + 
               geom_jitter(alpha = 0.1 ) +
               geom_boxplot(alpha = 0.2, color = 'red' )
```
```{r echo=FALSE, message=FALSE, warning=FALSE}
#plot barchart for density
ggplot(data = wine, aes(x = density)) +
  geom_histogram(binwidth = 0.001, color = 'black',fill = I('red'))
```

**INSIGHT:** Density seems to have a normal Distribution.

```{r echo=FALSE, message=FALSE, warning=FALSE}
#plot boxplot for pH
ggplot(wine, aes( x = 1, y = pH)) + 
               geom_jitter(alpha = 0.1 ) +
               geom_boxplot(alpha = 0.2, color = 'red' )
```

```{r echo=FALSE, message=FALSE, warning=FALSE}
#plot barchart for pH
ggplot(data = wine, aes(x = pH)) +
  geom_histogram(binwidth = 0.1, color = 'black',fill = I('red'))
```

**INSIGHT:** pH seems to have a normal Distribution.

```{r echo=FALSE, message=FALSE, warning=FALSE}
#plot boxplot for sulphates
ggplot(wine, aes( x = 1, y = sulphates)) + 
               geom_jitter(alpha = 0.1 ) +
               geom_boxplot(alpha = 0.2, color = 'red' ) +
               scale_y_continuous(lim = c(0.3,1.6))
```

```{r echo=FALSE, message=FALSE, warning=FALSE}
#plot barchart for sulphates
ggplot(data = wine, aes(x = sulphates)) +
  geom_histogram(binwidth = 0.1, color = 'black',fill = I('red')) +
  scale_x_continuous(lim = c(0.3,1.6))
```

**INSIGHT:** Sulphates has long tailed distribution.

```{r echo=FALSE, message=FALSE, warning=FALSE}
#plot boxplot for alcohol
ggplot(wine, aes( x = 1, y = alcohol)) + 
               geom_jitter(alpha = 0.1 ) +
               geom_boxplot(alpha = 0.2, color = 'red' ) +
               scale_y_continuous(lim = c(8,14))
```

```{r echo=FALSE, message=FALSE, warning=FALSE}
#plot barchart for alcohol
ggplot(data = wine, aes(x = alcohol)) +
  geom_histogram(binwidth = 0.1, color = 'black',fill = I('red')) +
  scale_x_continuous(lim = c(8,14))
```


**INSIGHT:** Alcohol has a skewed distribution.

## Analysis of the Univariate Plots

### What is the structure of your dataset?

This Dataset has 1599 rows and 14 columns.

Categorical variables are 'quality' and 'rating'. The rest of the variables are 
numerical variables which detail the chemical and physical properties of the 
wine.

### What is/are the main feature(s) of interest in your dataset?

My main focus in this dataset project is the 'quality' of the wine. What is it 
excalty that drives the quailty of the wine.

### What other features in the dataset do you think will help support your 
investigation into your feature(s) of interest?

After looking at the distributions of the varible I think the acidity of the 
wine is the biggest factor of it's quailty. Since pH is related to acidity, I 
also think it may have some effect on the quality. Residual sugar might also 
have an effect on wine quality, but from the little I know of wine, sugars 
should not have much of an effect on quality but on whether how 'dry' it is.

### Did you create any new variables from existing variables in the dataset?

I did do some modifications to the dataset. Quality seems like it would best fit
as a ordered factor so I converted it. I also added a new ordered factored 
variable called 'rating' based on the quailty of the wine.

### Of the features you investigated, were there any unusual distributions?

Residual sugar and Chloride seems to have extreme outliers present. Density and 
pH looks to be normally distributed with few outliers. Total and free sulfur 
dioxides, alcohol and sulphates, and Fixed and volatile acidity seem to be 
long-tailed. What was unusual was Citric acid. It has large number of zero 
values! It also showed a unique distribution; a mostly rectangular shape. 


## Bivariate Plots

For this section I think a correlation matrix should be created to start off. 
This matrix should give me an idea about which varibles may be correlated.

First let's remove some columns and convert quality back to a number!
```{r}
#remove some columns and convert quality back to a number
c <- cor(
  wine %>%
    dplyr::select(-X) %>%
    dplyr::select(-rating) %>%
    mutate(
      quality = as.numeric(quality)
    )
)
#create correlation matrix
emphasize.strong.cells(which(abs(c) > .3 & c != 1, arr.ind = TRUE))
pandoc.table(c)
```

**INSIGHT:** Quality is strongly correlated to Alcohol and Volatile acidity.

**INSIGHT:** Density and Fixed Acidity have a very strong correlation.

**INSIGHT:** Naturally, Alcohol has negative correlation with density.

**INSIGHT:** Very unsual! Volatile acidity has a positive correlation with pH...
I thought pH and acidity had a negative relationship! 

It seems like Box plots between these variables are in order!

```{r echo=FALSE, message=FALSE, warning=FALSE}
#plot boxplot of x = quality, y = fixed.acidity
ggplot(data = wine, aes(x = quality, y = fixed.acidity)) +
  geom_jitter( alpha = .25) +
  geom_boxplot(alpha = .25,color = 'green') +
  stat_summary(fun.y = "mean", 
               geom = "point", 
               color = "blue", 
               shape = 3, 
               size = 5)
```

**INSIGHT:** Fixed Acidity has almost no effect on the Quality.

```{r echo=FALSE, message=FALSE, warning=FALSE}
#plot boxplot of x = quality, y = volatile.acidity
ggplot(data=wine, aes(x = quality, y = volatile.acidity)) +
  geom_jitter( alpha = .25) +
  geom_boxplot(alpha = .25,color = 'green') +
  stat_summary(fun.y = "mean", 
               geom = "point", 
               color = "blue", 
               shape = 3, 
               size = 5)
```

**INSIGHT:** Volatile acid and quality seem to have a negative coorelation. 

```{r echo=FALSE, message=FALSE, warning=FALSE}
#plot boxplot of x=quality, y=citric.acid
ggplot(data=wine, aes(x=quality, y=citric.acid)) +
  geom_jitter( alpha = .25) +
  geom_boxplot(alpha = .25,color = 'green') +
  stat_summary(fun.y = "mean", 
               geom = "point", 
               color = "blue", 
               shape = 3, 
               size = 5)
```

**INSIGHT:** Citric acid and quailty seem to have a positive correlation.

```{r echo=FALSE, message=FALSE, warning=FALSE}
#plot boxplot of x=quality, y=residual.sugar
ggplot(data=wine, aes(x=quality, y=residual.sugar)) +
  geom_jitter( alpha = .25) +
  geom_boxplot(alpha = .25,color = 'green') +
  scale_y_continuous(lim = c(0,5)) +
  stat_summary(fun.y = "mean", 
               geom = "point", 
               color = "blue", 
               shape = 3, 
               size = 5)
```

**INSIGHT:** My previous thought that Residual Sugar may have no effect an
effect on the wine quality seems to be true!

```{r echo=FALSE, message=FALSE, warning=FALSE}
#plot boxplot of x=quality, y=chlorides
ggplot(data=wine, aes(x=quality, y=chlorides)) +
  geom_jitter( alpha = .25) +
  geom_boxplot(alpha = .25,color = 'green') +
  scale_y_continuous(lim = c(0,0.2)) +
  stat_summary(fun.y = "mean", 
               geom = "point", 
               color = "blue", 
               shape = 3, 
               size = 5)
```

**INSIGHT:** There seems to be a weak negative correalation between chlorides
and quality. 

```{r echo=FALSE, message=FALSE, warning=FALSE}
#plot boxplot of x=quality, y=free.sulfur.dioxide
ggplot(data=wine, aes(x=quality, y=free.sulfur.dioxide)) +
  geom_jitter( alpha = .25) +
  geom_boxplot(alpha = .25,color = 'green') +
  scale_y_continuous(lim = c(0,40)) +
  stat_summary(fun.y = "mean", 
               geom = "point", 
               color = "blue", 
               shape = 3, 
               size = 5)
```

**INSIGHT:** Very low concentrations of Free Sulphur Dioxide produces low
quality wine. And very high concentrations produces average wine.

```{r echo=FALSE, message=FALSE, warning=FALSE}
#plot boxplot of x=quality, y=total.sulfur.dioxide
ggplot(data=wine, aes(x=quality, y=total.sulfur.dioxide)) +
  geom_jitter( alpha = .25) +
  geom_boxplot(alpha = .25,color = 'green') +
  scale_y_continuous(lim = c(0,150)) +
  stat_summary(fun.y = "mean", 
               geom = "point", 
               color = "blue", 
               shape = 3, 
               size = 5)
```


**INSIGHT:** Very low concentrations of total Sulphur Dioxide produces low
quality wine. And very high concentrations produces average wine.

```{r echo=FALSE, message=FALSE, warning=FALSE}
#plot boxplot of x=quality, y=density
ggplot(data=wine, aes(x=quality, y=density)) +
  geom_jitter( alpha = .25) +
  geom_boxplot(alpha = .25,color = 'green') +
  stat_summary(fun.y = "mean", 
               geom = "point", 
               color = "blue", 
               shape = 3, 
               size = 5)
```


**INSIGHT:** Lower densities seem to favor higher quailty wines.

```{r echo=FALSE, message=FALSE, warning=FALSE}
#plot boxplot of x=quality, y=pH
ggplot(data=wine, aes(x=quality, y=pH)) +
  geom_jitter( alpha = .25) +
  geom_boxplot(alpha = .25,color = 'green') +
  stat_summary(fun.y = "mean", 
               geom = "point", 
               color = "blue", 
               shape = 3, 
               size = 5)
```


**INSIGHT:** Higher Quailty wines seem to contain less pH

```{r echo=FALSE, message=FALSE, warning=FALSE}
#plot scatterplot of x = fixed.acidity, y = pH
ggplot(data = wine, aes(x = fixed.acidity, y = pH)) +
  geom_point(alpha = 0.25) +
  scale_x_log10(breaks=seq(1,20,1)) +
  xlab("Fixed Acidity (Log Base 10)") +
  geom_smooth(method="lm")
```

```{r echo=FALSE, message=FALSE, warning=FALSE}
#plot scatterplot of x = volatile.acidity, y = pH
ggplot(data = wine, aes(x = volatile.acidity, y = pH)) +
  geom_point(alpha = 0.25) +
  scale_x_log10(breaks=seq(.1,1,.1)) +
  xlab("Volatile Acidity (Log Base 10)") +
  geom_smooth(method="lm")
```


```{r echo=FALSE, message=FALSE, warning=FALSE}
#plot scatterplot of x = citric.acid, y = pH
ggplot(data = subset(wine, citric.acid > 0), aes(x = citric.acid, y = pH)) +
  geom_point(alpha = 0.25) +
  scale_x_log10() +
  xlab("Citric Acid (Log Base 10)") +
  geom_smooth(method="lm")
```

**INSIGHT:** Why are we seeing that Volatile Acid and pH have a positive 
correlation? Let's pose this for further investigation.


```{r echo=FALSE, message=FALSE, warning=FALSE}
#plot boxplot of x=quality, y=sulphates
ggplot(data=wine, aes(x=quality, y=sulphates)) +
  geom_jitter( alpha = .25) +
  geom_boxplot(alpha = .25,color = 'green') +
  scale_y_continuous(lim = c(0.25,1)) +
  stat_summary(fun.y = "mean", 
               geom = "point", 
               color = "blue", 
               shape = 3, 
               size = 5)
```


**INSIGHT:** Even though we see many outliers in the 'Average' quality wine, it 
seems that better wines have a stronger concentration of Sulphates.

```{r echo=FALSE, message=FALSE, warning=FALSE}
#plot boxplot of x=quality, y=alcohol
ggplot(data=wine, aes(x=quality, y=alcohol)) +
  geom_jitter( alpha = .25) +
  geom_boxplot(alpha = .25,color = 'green') +
  stat_summary(fun.y = "mean", 
               geom = "point", 
               color = "blue", 
               shape = 3, 
               size = 5)
```

**INSIGHT:** Higher quality wines seem to have higher Alcohol content in it.
Below are some statitics to help clarify.

```{r echo=FALSE, message=FALSE, warning=FALSE}
#create quality/alcohol model
alc_qual_mod <- lm(as.numeric(quality) ~ alcohol,
                       data = wine)
summary(alc_qual_mod)
```

**INSIGHT:** The value of R squared seems to suggest that Alcohol contributes to
only about 23% of the quality. What about the other 77%? Let continue to dig. 
Let's do a correlation test quailty against each variable.

```{r echo=FALSE, message=FALSE, warning=FALSE}
#perform correlation tests
corr_test <- function(x, y) {
  return(cor.test(x, as.numeric(y))$estimate)
}
correlation_values <- c(
  corr_test(wine$fixed.acidity, wine$quality),
  corr_test(wine$volatile.acidity, wine$quality),
  corr_test(wine$citric.acid, wine$quality),
  corr_test(log10(wine$residual.sugar), wine$quality),
  corr_test(log10(wine$chlorides), wine$quality),
  corr_test(wine$free.sulfur.dioxide, wine$quality),
  corr_test(wine$total.sulfur.dioxide, wine$quality),
  corr_test(wine$density, wine$quality),
  corr_test(wine$pH, wine$quality),
  corr_test(log10(wine$sulphates), wine$quality),
  corr_test(wine$alcohol, wine$quality))
names(correlation_values) <- c('fixed.acidity', 'volatile.acidity', 'citric.acid',
                         'log10.residual.sugar',
                         'log10.chlordies', 'free.sulfur.dioxide',
                         'total.sulfur.dioxide', 'density', 'pH',
                         'log10.sulphates', 'alcohol')
correlation_values
```

**INSIGHT:** The correlation test shows that the following variables have a high
correlation to quality.

 1. Alcohol
 2. Sulphates(log10)
 3. Volatile Acidity
 4. Citric Acid

## Analysis of Bivariate Plots

### Talk about some of the relationships you observed in this part of the 
investigation. How did the feature(s) of interest vary with other features in 
the dataset? What was the strongest relationship you found?

Higher concentration of Citric Acid seems to produce higher quality. Higher 
alcohol percentages seems to produce higher quality. Lower percent of Chloride 
seems to produce higher quality. Lower densities seems to produce higher 
quality. The more acidic seems to produce higher quality. Lower Volatile Acidity
seems to produce higher quality. Residual sugar seems to have almost no effect 
on quality. Fixed Acidity seems to have almost no effect on quality.

### Did you observe any interesting relationships between the other features 
(not the main feature(s) of interest)?

For some reason volatile acidity has a positive correlation with pH...

## Multivariate Plots

The data shows that alcohol plays a strong part in the quality of the wine, so 
will make alcohol a constant and apply a few more variables to see if it effects
quality.

```{r echo=FALSE, message=FALSE, warning=FALSE}
# plot scatterplot of y = density, x = alcohol by quality
ggplot(data = wine,
       aes(y = density, x = alcohol,
           color = quality)) +
  geom_point(alpha = 0.5, size = 1) +
  geom_smooth(method = "lm", se = FALSE,size=1)  +
  scale_color_brewer(type='seq',
                   guide=guide_legend(title='Quality Value'))
```

**INSIGHT:** The quality of the alcohol does seem be effected by Density. 

```{r echo=FALSE, message=FALSE, warning=FALSE}
# plot scatterplot of y = sulphates, x = alcohol by quality
ggplot(data = wine,
       aes(y = sulphates, x = alcohol,
           color = quality)) +
  geom_point(alpha = 0.5, size = 1) +
  geom_smooth(method = "lm", se = FALSE,size=1)  +
  scale_y_continuous(limits=c(0.3,1.6)) +
  facet_wrap(~rating) +
  scale_color_brewer(type='seq',
                   guide=guide_legend(title='Quality Value'))
```


**INSIGHT:** It seems like a high the level of Sulphates paired with higher
alcohol content produce higher quality.

```{r echo=FALSE, message=FALSE, warning=FALSE}
# plot scatterplot of y = volatile.acidity x = alcohol by quality
ggplot(data = wine,
       aes(y = volatile.acidity, x = alcohol,
           color = quality)) +
  geom_point(alpha = 0.5, size = 1) +
  geom_smooth(method = "lm", se = FALSE,size=1)  +
  facet_wrap(~rating) +
  scale_color_brewer(type='seq',
                   guide=guide_legend(title='Quality Value'))
```

**INSIGHT:** It seems like a low the level of Volatile acid paired with higher 
alcohol content produce higher quality.

```{r echo=FALSE, message=FALSE, warning=FALSE}
# plot scatterplot of y = pH x = alcohol by quality
ggplot(data = wine,
       aes(y = pH, x = alcohol,
           color = quality)) +
  geom_point(alpha = 0.5, size = 1) +
  geom_smooth(method = "lm", se = FALSE,size=1)  +
  facet_wrap(~rating) +
  scale_color_brewer(type='seq',
                   guide=guide_legend(title='Quality Value'))
```


**INSIGHT:** It seems like a low pH level paired with higher alcohol content 
produce higher quality.

```{r echo=FALSE, message=FALSE, warning=FALSE}
# plot scatterplot of y = residual.sugar x = alcohol by quality
ggplot(data = wine,
       aes(y = residual.sugar, x = alcohol,
           color = quality)) +
  geom_point(alpha = 0.5, size = 1) +
  geom_smooth(method = "lm", se = FALSE,size=1)  +
  facet_wrap(~rating) +
  scale_color_brewer(type='seq',
                   guide=guide_legend(title='Quality Value'))
```

**INSIGHT:** No correlation between residual sugar and quality keeping alcohol 
constant.

```{r echo=FALSE, message=FALSE, warning=FALSE}
# plot scatterplot of y = total.sulfur.dioxide x = alcohol by quality
ggplot(data = wine,
       aes(y = total.sulfur.dioxide, x = alcohol,
           color = quality)) +
  geom_point(alpha = 0.5, size = 1) +
  geom_smooth(method = "lm", se = FALSE,size=1)  +
  facet_wrap(~rating) +
  scale_color_brewer(type='seq',
                   guide=guide_legend(title='Quality Value'))
```


**INSIGHT:** Lower Sulphur Dioxide seems to produces better wine

Now we will see the effect of Acids on the Quality.

```{r echo=FALSE, message=FALSE, warning=FALSE}
# plot scatterplot of y = citric.acid, x = volatile.acidity by quality
ggplot(data = wine,
       aes(y = citric.acid, x = volatile.acidity,
           color = quality)) +
  geom_point(alpha = 0.5, size = 1) +
  geom_smooth(method = "lm", se = FALSE,size=1)  +
  facet_wrap(~rating) +
  scale_color_brewer(type='seq',
                   guide=guide_legend(title='Quality Value'))
```

**INSIGHT:** Higher Citric Acid and low Volatile Acid seems to produce better
Wines

```{r echo=FALSE, message=FALSE, warning=FALSE}
# plot scatterplot of y = citric.acid, x = fixed.acidity by quality
ggplot(data = wine,
       aes(y = citric.acid, x = fixed.acidity,
           color = quality)) +
  geom_point(alpha = 0.5, size = 1) +
  geom_smooth(method = "lm", se = FALSE,size=1)  +
  facet_wrap(~rating) +
  scale_color_brewer(type='seq',
                   guide=guide_legend(title='Quality Value'))
```

**INSIGHT:** Nothing here.

```{r echo=FALSE, message=FALSE, warning=FALSE}
# plot scatterplot of fixed.acidity, x = volatile.acidity by quality
ggplot(data = wine,
       aes(y = fixed.acidity, x = volatile.acidity,
           color = quality)) +
  geom_point(alpha = 0.5, size = 1) +
  geom_smooth(method = "lm", se = FALSE,size=1)  +
  facet_wrap(~rating) +
  scale_color_brewer(type='seq',
                   guide=guide_legend(title='Quality Value'))
```


**INSIGHT:** Nothing here.

## Linear Modelling

Now after all these analysis, I am going to take the variables which are most 
strongly correlated with the quality of the wine and generate a linear model 
with them.

```{r echo=FALSE, message=FALSE, warning=FALSE}
# create linear model
set.seed(1221)
tr_data <- sample_frac(wine, .6)
te_data <- wine[ !wine$X %in% tr_data$X, ]
m1 <- lm(as.numeric(quality) ~ alcohol, data = tr_data)
m2 <- update(m1, ~ . + sulphates)
m3 <- update(m2, ~ . + volatile.acidity)
m4 <- update(m3, ~ . + citric.acid)
m5 <- update(m4, ~ . + fixed.acidity)
m6 <- update(m2, ~ . + pH)
mtable(m1,m2,m3,m4,m5,m6)
```


```{r echo=FALSE, message=FALSE, warning=FALSE}
#execute model and predict values
qual_predict <- data.frame(
  te_data$quality,
  predict(m5, te_data) - as.numeric(te_data$quality)
)
names(qual_predict) <- c("quality", "error")
#plot scatterplot of errors vs quality
ggplot(data=qual_predict, aes(x=quality,y=error)) +
  geom_jitter(alpha = 0.3)
```

## Analysis of the Multivariate Plots

### Talk about some of the relationships you observed in this part of the 
investigation. Were there features that strengthened each other in terms of 
looking at your feature(s) of interest? Were there any interesting or surprising
interactions between features?

High Alcohol coupled with high Sulaphate seems to produce higher quality. High 
Alcohol coupled with Citric Acid seems to show a weak positive correlation for 
higher quality.


### OPTIONAL: Did you create any models with your dataset? Discuss the strengths
and limitations of your model.

Using the data insights gathered, a linear model was created. But the model did 
suffer from a limitation. Which was that the majority of the data was for 
average quality wines. Therefore a significant confidence level in the equations
was not produced. This is explained by the low R squared value obtained; alcohol
only contributes to only 23% of the Quality. A more complete dataset with more 
'good' and 'bad' quality wines is needed.

## Final Plots and Summary

1- Alcohol and Sulphates seems to be important in producing higher wine quality.

2- The linear model showed us the variation in the error percentage with 
different qualities of Wine, so plot of error value against the quality should 
useful.

### Plot 1

```{r echo=FALSE, message=FALSE, warning=FALSE}
#create boxplot of y=alcohol, x=quality
ggplot(data=wine, aes(y=alcohol, x=quality)) + 
  geom_jitter(alpha = .25)  +
  geom_boxplot(alpha = .25,color = 'green') +
  stat_summary(fun.y = "mean", 
               geom = "point", 
               color = "blue", 
               shape = 3, 
               size = 5) +
  xlab("Quality") +
  ggtitle("Effect of alcohol on wine quality")
```

**INSIGHT:** It's easy to see that Alcohol has a positive relationship with 
quality. The more the alcohol is present, the higher the quality. But the R 
Squared value shows that alcohol is only responsible for about 23% in the 
variance of the quality. 

### Plot 2

```{r echo=FALSE, message=FALSE, warning=FALSE}
#create scatterplot of y = sulphates x = alcohol by quality
ggplot(data = wine,
       aes(y = sulphates, x = alcohol,
           color = quality)) +
  geom_point(alpha = 0.5, size = 1) +
  geom_smooth(method = "lm", se = FALSE,size=1)  +
  scale_y_continuous(limits=c(0.3,1.5)) +
  ylab("potassium sulphate (g/dm3)") +
  xlab("Alcohol Percent") +
  scale_color_brewer(type='seq',
                   guide=guide_legend(title='Quality')) +
  ggtitle("Alcohol + sulphates effects wine quality")
```



**INSIGHT:** This plot shows that wines with both high values for Alcohol 
percentage and Sulphates seems to yield higher quality wine.

### Plot 3


```{r echo=FALSE, message=FALSE, warning=FALSE }
#plot scatterplot of errors vs quality
df_1 <- data.frame(
  te_data$quality,
  predict(m5, te_data) - as.numeric(te_data$quality)
)
names(df_1) <- c("quality", "error")
ggplot(data=df_1, aes(x=quality,y=error)) +
  geom_jitter(alpha = 0.25) +
  ggtitle("Errors vs expected quality")
```


**INSIGHT:** Since most of the data is for 'Average' quality wines, it 
makes sense that the error is super dense around that section. m5, the best fit 
linear model with the highest R squared value could only explain 32% change in 
quality. So to summarize, predicting both 'Good' and 'Bad' quality wines using 
this model is ill advised.


## Reflections

Since most of the data collected was for 'average' quality, the training set was
not helpful in building a model that can  accurately predict the quality of wine
in the extreme ranges ('bad' and 'good' wines). A data set with more infomation 
on bad and good quality wines is needed.

My univariate analysis, showed that most variables displayed either a Positively
skewed or a Normal Distribution. But the distribution of Citric acid looked a 
bit off, some wines samples had no citric acid present, seems that there is 
incomplete data present.

My bivariate analysis, plotting variables vs quality, showed that the most 
influential factors for quality were alcohol, sulphate and acid concentrations. 
Also a very weird relationship showed up, Volatile acidity had a positive 
correlation with pH... I thought pH and acidity had a negative relationship! 

My multivariate analysis, showed that Alcohol has a positive relationship with 
quality. The more the alcohol is present, the higher the quality. But the 
R Squared value shows that alcohol is only responsible for about 23% in the 
variance of the quality. Density seemed to have no part in wine quality. And 
that high values for Alcohol percentage and Sulphates seems to yield higher 
quality wine.

Suggestions for the future: Maybe a more complete dataset should be collected. 
One with more values in the good and bad ranges. Also, Volatile acidity having a
positive correlation with pH is super weird. Maybe an in-depth analysis of this 
should done.
